---
title: "Untargeted Metabolomics Analysis Report"
subtitle: "Southeast Center for Integrated Metabolomics (SECIM)"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    embed-resources: true
    smooth-scroll: true
    fig-width: 10
    fig-height: 8
params:
  client: "ExampleClient"
  PI_name: "Dr. Jane Smith"
  test_type: "t.test"
  contrast_var: "Class"
  input_file: "InputFiles/example.xlsx"
  data_file: "Results/ExampleClient/ExampleClient_ReportData.qs"
  processing_time: 10.5
---

```{r setup}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

```{r load-libraries}
#| label: load-libraries
#| include: false

suppressPackageStartupMessages({
  library(qs)
  library(dplyr)
  library(ggplot2)
  library(plotly)
  library(DT)
  library(stringr)
  library(RColorBrewer)
  library(heatmaply)
  library(gridExtra)
  library(grid)
  library(car)
  library(htmltools)
  library(downloadthis)
})
```

```{r load-data}
#| label: load-data
#| include: false

# Load serialized results from Stage 1
if (!file.exists(params$data_file)) {
  stop(sprintf("Data file not found: %s", params$data_file))
}

ReportInput <- qs::qread(params$data_file)
Client_Data_Download <- ReportInput

# Extract parameters
client <- params$client
PI_name <- params$PI_name
test_type <- params$test_type
contrast_var <- params$contrast_var
```

```{r process-data}
#| label: process-data
#| include: false

# Convert confidence levels: 1 → High, 3 → Low
modify_data_frame <- function(df) {
  if ("Level" %in% colnames(df)) {
    colnames(df)[colnames(df) == "Level"] <- "ID_confidence"
    df$ID_confidence <- as.character(df$ID_confidence)
    df$ID_confidence[df$ID_confidence == "1"] <- "High"
    df$ID_confidence[df$ID_confidence == "3"] <- "Low"
  }
  return(df)
}

Client_Data_Download <- lapply(Client_Data_Download, function(x) {
  if (is.data.frame(x)) modify_data_frame(x) else x
})

# Remove "Zref_" prefixes
clean_data <- function(df) {
  colnames(df) <- gsub("Zref_", "", colnames(df))
  df[] <- lapply(df, function(x) {
    if (is.character(x)) gsub("Zref_", "", x)
    else if (is.factor(x)) { levels(x) <- gsub("Zref_", "", levels(x)); x }
    else x
  })
  return(df)
}

for (i in seq_along(Client_Data_Download)) {
  if (is.data.frame(Client_Data_Download[[i]])) {
    Client_Data_Download[[i]] <- clean_data(Client_Data_Download[[i]])
  }
}

# Determine significance threshold
if (test_type %in% c("lm", "lme", "anova", "t.test")) {
  p_type <- ifelse(
    nrow(Client_Data_Download[["report_results"]] %>% filter(adj.p.value < 0.05)) > 10,
    "adj.p.value", "p.value"
  )
  sig_knowns <- nrow(
    Client_Data_Download[["report_results"]] %>%
      filter(!!sym(p_type) < 0.05, ID_confidence == "High")
  )
} else if (test_type == "nostats") {
  p_type <- "log2FC"
  sig_knowns <- nrow(
    Client_Data_Download[["report_results"]] %>%
      filter(abs(log2FC) > 0.58, ID_confidence == "High")
  )
}

contrasts <- levels(as.factor(Client_Data_Download[["report_results"]]$contrast))
```

# Project Summary

**Principal Investigator:** `r PI_name`  
**Client:** `r client`  
**Analysis Date:** `r format(Sys.time(), '%B %d, %Y')`  
**Processing Time:** `r sprintf("%.2f minutes", params$processing_time)`

## Sample Information

```{r sample-table}
#| label: sample-table

datatable(
  data.frame(table(Client_Data_Download$metadata[[contrast_var]])),
  colnames = c(contrast_var, "N"),
  options = list(dom = 't', paging = FALSE, searching = FALSE, ordering = FALSE),
  rownames = FALSE
)
```

## Download Results

```{r download-button}
#| label: download-button

# Prepare Excel download
Client_Data_Download_out <- Client_Data_Download[!grepl("View", names(Client_Data_Download))]

Client_Data_Download_out %>%
  download_this(
    output_name = paste0(client, "_Results"),
    output_extension = ".xlsx",
    button_label = "Download Complete Results",
    button_type = "primary",
    has_icon = TRUE,
    icon = "fa fa-download"
  )
```

# Methods

## Instrumental Analysis

Global metabolomics profiling was performed on a **Thermo Q-Exactive Orbitrap** mass spectrometer with Dionex UHPLC.

- **Column:** ACE 18-PFP (100 × 2.1 mm, 2 µm)
- **Mobile Phases:** A = 0.1% formic acid in water; B = acetonitrile
- **Flow Rate:** 350 µL/min at 25°C
- **Ionization:** Positive and negative heated electrospray
- **Resolution:** 35,000 at m/z 200

## Data Processing & Statistical Analysis

```{r methods-text}
#| label: methods-text
#| results: asis

cat("**Feature Detection:** MZmine was used for peak detection, deisotoping, and alignment.\n\n")
cat("**Quality Control:** Blank feature filtering (IQR method) and KNN imputation via MetaboAnalystR.\n\n")
cat("**Normalization:** Sum normalization, log transformation, and Pareto scaling.\n\n")

if (test_type == "t.test") {
  cat("**Statistics:** Welch's t-test with FDR correction (Benjamini-Hochberg).\n")
} else if (test_type == "anova") {
  cat("**Statistics:** ANOVA with post-hoc pairwise comparisons via estimated marginal means (EMMs).\n")
} else if (test_type %in% c("lm", "lme")) {
  cat("**Statistics:** Linear mixed model with post-hoc EMMs comparisons.\n")
} else {
  cat("**Analysis:** Fold-change analysis (|log2FC| > 1.5 threshold).\n")
}
```

# Results

## Principal Component Analysis

```{r pca-plot}
#| label: pca-plot
#| fig-width: 10
#| fig-height: 6

# Combine normalized data
PCA_DATA <- t(rbind(
  Client_Data_Download[["Neg.normalized.data"]][, 3:(ncol(Client_Data_Download[["Neg.normalized.data"]]) - 2)],
  Client_Data_Download[["Pos.normalized.data"]][, 3:(ncol(Client_Data_Download[["Pos.normalized.data"]]) - 2)]
))

# Remove metadata rows
metadata_rows <- which(is.na(as.numeric(PCA_DATA[, 1])))
if (length(metadata_rows) > 0) PCA_DATA <- PCA_DATA[-metadata_rows, ]

# Perform PCA
numsonly <- data.frame(lapply(as.data.frame(PCA_DATA[, -1]), as.numeric))
pca_res <- prcomp(numsonly, center = TRUE, scale. = FALSE, rank. = 2)
scores <- as.data.frame(pca_res$x)

# Add metadata
scores$Sample <- rownames(PCA_DATA)
scores$Group <- Client_Data_Download$metadata[[contrast_var]][match(scores$Sample, rownames(Client_Data_Download$metadata))]

# Interactive PCA plot
colors <- RColorBrewer::brewer.pal(max(3, length(unique(scores$Group))), "Set1")

fig <- plot_ly()

# Add ellipses and points per group
for (group in unique(scores$Group)) {
  group_data <- scores[scores$Group == group, ]
  group_color <- colors[which(unique(scores$Group) == group)]
  
  # Ellipse
  if (nrow(group_data) > 2) {
    ellipse_coords <- car::dataEllipse(
      group_data$PC1, group_data$PC2,
      levels = 0.68, plot.points = FALSE, draw = FALSE
    )
    
    fig <- fig %>%
      add_polygons(
        x = ellipse_coords[, 1], y = ellipse_coords[, 2],
        line = list(color = group_color, dash = "dot"),
        fillcolor = group_color, opacity = 0.3,
        showlegend = FALSE, hoverinfo = "skip"
      )
  }
  
  # Points
  fig <- fig %>%
    add_trace(
      data = group_data,
      x = ~PC1, y = ~PC2,
      type = "scatter", mode = "markers",
      name = group,
      marker = list(color = group_color, size = 10),
      text = ~Sample,
      hovertemplate = paste0("<b>", group, "</b><br>%{text}<br>PC1: %{x:.2f}<br>PC2: %{y:.2f}<extra></extra>")
    )
}

fig <- fig %>%
  layout(
    title = "Principal Component Analysis",
    xaxis = list(title = sprintf("PC1 (%.1f%% variance)", summary(pca_res)$importance[2, 1] * 100)),
    yaxis = list(title = sprintf("PC2 (%.1f%% variance)", summary(pca_res)$importance[2, 2] * 100)),
    hovermode = "closest"
  )

fig
```

## Significantly Changed Metabolites

```{r sig-metabolites-table}
#| label: sig-metabolites-table

if (p_type == "log2FC") {
  results_filtered <- Client_Data_Download[["report_results"]] %>%
    filter(abs(log2FC) > 1.5) %>%
    select(compound, ID_confidence, contrast, FC, log2FC) %>%
    arrange(desc(abs(log2FC)))
  
  caption_text <- "Changed metabolites (|log2FC| > 1.5)"
} else {
  results_filtered <- Client_Data_Download[["report_results"]] %>%
    filter(!!sym(p_type) < 0.05) %>%
    select(compound, !!sym(p_type), ID_confidence, contrast, FC, log2FC) %>%
    arrange(!!sym(p_type))
  
  caption_text <- sprintf("Significantly changed metabolites (%s < 0.05)", p_type)
}

datatable(
  results_filtered,
  caption = caption_text,
  filter = "top",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  extensions = 'Buttons'
) %>%
  formatRound(columns = c("FC", "log2FC"), digits = 3)
```

## Volcano Plots {.tabset}

::: {.callout-tip}
## Interactive Features
- Hover over points to view metabolite names
- **Red:** Significant + large fold-change
- **Orange:** Significant only
- **Black:** Large fold-change only
- **Gray:** Not significant
:::

```{r volcano-plots}
#| label: volcano-plots
#| results: asis

if (test_type != "nostats") {
  for (i in seq_along(contrasts)) {
    cat(sprintf("\n### %s\n\n", contrasts[i]))
    
    volcano_data <- Client_Data_Download[["report_results"]] %>%
      filter(contrast == contrasts[i])
    
    # Assign groups
    volcano_data$group <- "NotSignificant"
    volcano_data$group[volcano_data[[p_type]] < 0.05 & abs(volcano_data$log2FC) < 0.58] <- "Significant"
    volcano_data$group[abs(volcano_data$log2FC) > 0.58] <- "log2FC"
    volcano_data$group[volcano_data[[p_type]] < 0.05 & abs(volcano_data$log2FC) > 0.58] <- "Significant&log2FC"
    
    # Create plot
    p <- plot_ly(
      data = volcano_data,
      x = ~log2FC,
      y = ~-log10(get(p_type)),
      color = ~group,
      colors = c(
        "NotSignificant" = "gray",
        "Significant" = "orange",
        "log2FC" = "black",
        "Significant&log2FC" = "red"
      ),
      text = ~paste0(compound, " (", ID_confidence, ")"),
      hovertemplate = "%{text}<br>log2FC: %{x:.2f}<br>-log10(p): %{y:.2f}<extra></extra>",
      type = "scatter",
      mode = "markers"
    ) %>%
      layout(
        title = sprintf("Volcano Plot: %s", contrasts[i]),
        xaxis = list(title = "log2(Fold Change)", zeroline = TRUE),
        yaxis = list(title = sprintf("-log10(%s)", p_type)),
        showlegend = TRUE
      )
    
    print(htmltools::tagList(p))
    cat("\n\n")
  }
}
```

## Normalization Quality Control {.tabset}

### Positive Ion Mode

```{r norm-pos}
#| label: norm-pos
#| fig-height: 10

grid.arrange(
  Client_Data_Download[["Pos.FeatureView"]]$grobs[[1]],
  Client_Data_Download[["Pos.FeatureView"]]$grobs[[2]],
  Client_Data_Download[["Pos.FeatureView"]]$grobs[[3]],
  Client_Data_Download[["Pos.FeatureView"]]$grobs[[4]],
  ncol = 2,
  top = textGrob("Positive Mode - Feature View", gp = gpar(fontsize = 14, font = 2))
)
```

### Negative Ion Mode

```{r norm-neg}
#| label: norm-neg
#| fig-height: 10

grid.arrange(
  Client_Data_Download[["Neg.FeatureView"]]$grobs[[1]],
  Client_Data_Download[["Neg.FeatureView"]]$grobs[[2]],
  Client_Data_Download[["Neg.FeatureView"]]$grobs[[3]],
  Client_Data_Download[["Neg.FeatureView"]]$grobs[[4]],
  ncol = 2,
  top = textGrob("Negative Mode - Feature View", gp = gpar(fontsize = 14, font = 2))
)
```

## Top Changed Metabolites Heatmap {.tabset}

```{r heatmaps}
#| label: heatmaps
#| results: asis
#| fig-height: 10

for (i in seq_along(contrasts)) {
  cat(sprintf("\n### %s\n\n", contrasts[i]))
  
  # Get top 20 metabolites
  if (p_type == "log2FC") {
    top_metabs <- Client_Data_Download[["report_results"]] %>%
      filter(contrast == contrasts[i], abs(log2FC) > 1.5) %>%
      arrange(desc(abs(log2FC))) %>%
      slice_head(n = 20) %>%
      pull(row.mode)
  } else {
    top_metabs <- Client_Data_Download[["report_results"]] %>%
      filter(contrast == contrasts[i], !!sym(p_type) < 0.05) %>%
      arrange(!!sym(p_type)) %>%
      slice_head(n = 20) %>%
      pull(row.mode)
  }
  
  if (length(top_metabs) > 0) {
    # Extract normalized data
    norm_data <- rbind(
      Client_Data_Download[["Pos.normalized.data"]],
      Client_Data_Download[["Neg.normalized.data"]]
    )
    
    norm_data$row.mode <- paste(norm_data$row.ID, norm_data$mode, sep = "_")
    heatmap_data <- norm_data %>%
      filter(row.mode %in% top_metabs) %>%
      select(-row.ID, -mode, -row.mode, -compound, -mz, -rt)
    
    rownames(heatmap_data) <- norm_data$compound[norm_data$row.mode %in% top_metabs]
    heatmap_data <- as.matrix(heatmap_data)
    
    # Create heatmap
    hm <- heatmaply(
      heatmap_data,
      scale = "row",
      colors = colorRampPalette(c("darkblue", "white", "darkred"))(100),
      showticklabels = c(FALSE, TRUE),
      ylab = "Metabolites",
      xlab = "Samples",
      main = sprintf("Top Metabolites: %s", contrasts[i]),
      fontsize_row = 10
    )
    
    print(htmltools::tagList(hm))
  } else {
    cat("*No significantly changed metabolites for this contrast.*\n")
  }
  
  cat("\n\n")
}
```

# Session Information

```{r session-info}
#| label: session-info

sessionInfo()
```
